---
title: "Diabetes and Healthcare Utilization Among Medicare Beneficiaries"
subtitle: "A claims-based comparative cohort study (CMS DE-SynPUF, synthetic data)"
author: "Gael Songwa"
date: today
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
    number-sections: true
    theme: cosmo
    code-fold: true
    df-print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(dplyr)
library(readr)
root <- rprojroot::find_root(rprojroot::is_rstudio_project)
cohort   <- readRDS(file.path(root, "data", "cohort.rds"))
co       <- readRDS(file.path(root, "data", "cohort_outcomes.rds"))
m_inp    <- readRDS(file.path(root, "data", "model_inpatient.rds"))
m_out    <- readRDS(file.path(root, "data", "model_outpatient.rds"))
```

# Background

Diabetes is associated with substantial healthcare use, but crude comparisons are
confounded by age and comorbidity. This study estimates the **independent**
association between diabetes and 2009 healthcare utilization in a Medicare cohort,
adjusting for measured confounders.

> **Data caveat.** This analysis uses CMS DE-SynPUF *synthetic* claims. It mirrors the
> structure of Medicare data but has no inferential validity; results are a methods
> demonstration, not clinical evidence.

# Methods

**Design.** Retrospective comparative cohort. Baseline year 2008; index date
2009-01-01; follow-up year 2009.

**Population.** Beneficiaries present in the 2008 Beneficiary Summary File, aged
≥ 65 at index, alive at index, and continuously enrolled (Part A & B, 12 months)
in 2008.

**Exposure.** Diabetes at baseline (`SP_DIABETES`).

**Outcomes.** Counts of 2009 inpatient admissions (primary) and outpatient visits
(secondary).

**Analysis.** Negative binomial regression (chosen over Poisson after confirming
overdispersion), adjusted for age, sex, and comorbidities (CHF, CKD, COPD, IHD).
Rate ratios (RR) reported with 95% confidence intervals.

# Results

## Cohort attrition

```{r attrition}
tibble::tribble(
  ~Step,                                   ~N,
  "Baseline population (2008 file)",       116352L,
  "After data-quality (birth date / sex)", 116352L,
  "After age >= 65",                       98110L,
  "After alive at index",                  96528L,
  "After continuous enrollment",           80988L
)
```

The final analytic cohort included **`r nrow(co)`** beneficiaries
(`r sum(co$has_diabetes)` with diabetes, `r sum(!co$has_diabetes)` without).

## Table 1: baseline characteristics by diabetes group

```{r table1}
co |>
  group_by(Diabetes = has_diabetes) |>
  summarise(
    N            = n(),
    `Age median` = median(age_at_index),
    `% female`   = round(mean(BENE_SEX_IDENT_CD == 2) * 100, 1),
    `% CHF`      = round(mean(SP_CHF == 1) * 100, 1),
    `% CKD`      = round(mean(SP_CHRNKIDN == 1) * 100, 1),
    `% COPD`     = round(mean(SP_COPD == 1) * 100, 1),
    `% IHD`      = round(mean(SP_ISCHMCHT == 1) * 100, 1),
    .groups = "drop"
  )
```

Diabetics were older and substantially more comorbid across all conditions,
motivating covariate adjustment.

## Model results: adjusted rate ratios

```{r results-inpatient}
round(m_inp$rr_table, 3)
```

*Inpatient admissions.* Adjusted RR for diabetes:
**`r round(m_inp$rr_table["has_diabetesTRUE", "RR"], 2)`**
(95% CI `r round(m_inp$rr_table["has_diabetesTRUE", "CI_low"], 2)`–
`r round(m_inp$rr_table["has_diabetesTRUE", "CI_high"], 2)`).

```{r results-outpatient}
round(m_out$rr_table, 3)
```

*Outpatient visits.* Adjusted RR for diabetes:
**`r round(m_out$rr_table["has_diabetesTRUE", "RR"], 2)`**
(95% CI `r round(m_out$rr_table["has_diabetesTRUE", "CI_low"], 2)`–
`r round(m_out$rr_table["has_diabetesTRUE", "CI_high"], 2)`).

# Interpretation

After adjustment, diabetes was independently associated with roughly two-thirds
more inpatient admissions and outpatient visits. The crude admission association
(RR ~3.1) shrank markedly after adjustment (RR ~1.6), illustrating substantial
confounding by age and comorbidity; the outpatient association shrank less,
consistent with comorbidities confounding hospitalizations more than routine visits.

**Limitations.** Synthetic data (no inferential validity); comorbidities may lie on
the causal pathway from diabetes, so adjustment may underestimate the total effect;
a fixed one-year window ignores differential follow-up time.

# Reproducibility

Environment pinned with `renv`; pipeline (`download → cohort → outcomes → models`)
and data-quality tests run in CI on every change.

